Supplementary Information

### 2020-11-29 doc/manuscript/MEE_manuscript.Rmd
manuscript work

#Figures
schematic=1
SCBI_example=2
clim_sens=3
size=4
full_examples = 5
# Fig5/ table2- DBH/climate interactions
year_fig = 7

@@

# Abstract

@@

@@

3. Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$. The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).

@@

# Introduction

@@

# Materials and Methods

## Data sources and preparation

@@

## Analysis methods

@@

### Identifying key climate drivers

@@

### Combining drivers in GLS model

@@

# Results

## Comparison with traditional methods

@@

## Identifying climate drivers

@@

## Climate sensitivity

@@

## Influence of DBH

@@

## Additive and interactive effects of climate and DBH

@@

**Interactive effects of climate and $DBH$ were found for # of the # species-variable combinations for which they could be tested**. 
Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r interactions_and_year_summaries`).
The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig. **6?**).
Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r interactions_and_year_summaries`). 
The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).

@@

## Effects of year

@@

# Discussion

**We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth (Fig. `r schematic`).**
Results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs. `r SCBI_example`, `r all_traditional_comparisons`,  Table `r qualitative_results_comparison`), but offer several new insights.
We show that ..
Most notably, the fact that our approach is able to simultaneously model the effects of climate, DBH, and year allows...

## climate sensitivity 
*corroboration of insight from conventional methods*
*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*, with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential

@@

## size trends 
*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*

## changing growth rates
- growth rates changing directionally
- stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.

@@

## future challenges (?)
*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*

@@

# Misc content/ notes to integrate
- Scotty temperture responses are probably not physiological, but due to permafrost.

# Acknowledgements
Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
Helpful feedback was provided by Helene Muller-Landau, Albert Kim...
This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S

# Authors' contributions
KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
[most authors] collected the data; 
VH, BG, EGA, and NP organized and analysed the data; 
KAT led the writing of the manuscript. 
All authors contributed critically to the drafts and gave final approval for publication.

# Data availability
```{r eval=FALSE, echo=FALSE}
# To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
```

@@

# Supplementary files

@@

# References

>>>>

#Figures
schematic=1
SCBI_example=2
clim_sens=3
size=4
full_examples = 5
interactions_fig = 6 #DBH/climate interactions
year_fig = 7

@@

# Abstract  {-#Abstract}

@@

*(needs to be re-written)*

@@

3. Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$. The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested). After accounting for size, growth rate varied directionally over time

@@

# Introduction {-#Introduction}

@@

# Materials and Methods {-#Materials}

## Data sources and preparation {-#Data}

@@

## Analysis methods {-#Analysis}

@@

### Identifying key climate drivers {-#Identifying}

@@

### Combining drivers in GLS model {-#Combining}

@@

# Results {-#Results}

## Comparison with traditional methods {-#Comparison}

@@

## Identifying climate drivers {-#Identifying}

@@

## Climate sensitivity {-#Climate}

@@

## Influence of DBH {-#Influence}

@@

## Additive and interactive effects of climate and DBH {-#Additive}

@@

**Interactive effects of climate and $DBH$ were found for # of the # species-variable combinations for which they could be tested**. 
Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r interactions_and_year_summaries`).
The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.  `r interactions_fig`).
Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r interactions_and_year_summaries`). 
The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig.  `r interactions_fig`).

@@

## Effects of year {-#Effects}

@@

# Discussion {-#Discussion}

**Our new tree-ring analysis method allows simultaneous consideration of the effects of objectively determined principle climate drivers, tree size, and year on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.** 
Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded
positively to water availability ($PPT$, $PDF$, or **$SF$**)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
Additive effects of water and temperature variables, usually over different time windows, occurred for most species, and nonlinear responses were common.
Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`), with larger trees tending to be more drought sensitive (REFS) *but less strongly affected by high temperatures (REFS)*.
Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ and was dependent upon species ecology in interaction with stand history and global change drivers (Figs. `r size`, `r year_fig`).
While none of these findings are biologically surprising, the simultaneous modeling of growth responses to climate, $DBH$, and year offers a fuller picture of the drivers of tree growth in an era of global change.

## Climate sensitivity {-#Climate}
*(corroboration of insight from conventional methods)*

Results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), but offer several new insights.

*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*, with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential

@@

## Size trends {-#Size}
*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*

## Changing growth rates {-#Changing}
- growth rates changing directionally
- stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.

@@

## Future challenges (?) {-#Future}
*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*

@@

# Misc content/ notes to integrate {-#Misc}
- Scotty temperture responses are probably not physiological, but due to permafrost.

# Acknowledgements {-#Acknowledgements}
Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
Helpful feedback was provided by Helene Muller-Landau, Albert Kim...
This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S

# Authors' contributions {-#Authors}
KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
[most authors] collected the data; 
VH, BG, EGA, and NP organized and analysed the data; 
KAT led the writing of the manuscript. 
All authors contributed critically to the drafts and gave final approval for publication.

# Data availability {-#Data}
```{r eval=FALSE, echo=FALSE}
# To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
```

@@

# Supplementary files {-#Supplementary}

@@

# References {-#References}
